---
title: "Specify a diagnostic model"
weight: 1
description: |
  Get started by learning how to specify and estimate a model using measr.
css: ../styles.css
link-citations: true
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

source(here::here("common.R"))
```

```{r}
#| label: load
#| include: false
#| message: false
#| warning: false

library(dcmdata)
library(dcmstan)
library(measr)
pkgs <- c("dcmdata", "measr", "rstan")
```

## Introduction

How do you specify and estimate a diagnostic classification model (DCM) using measr?
In this article, we will walk you through the steps.
We start with data for building the model, learn how to specify DCMs that make different assumptions about the data, and explore how to estimate the model with [*Stan*](https://mc-stan.org).

`r article_req_pkgs(pkgs)`

```{r}
#| label: user-load
#| message: false

library(measr)     # for model specification and estimation

# Helper packages
library(dcmdata)   # for example data set
library(rstan)     # for estimation backend
```

## Rapid Online Assessment of Reading and Phonological Awareness (ROAR-PA) data

Let's use data from the ROAR-PA [@roarpa] to learn how to specify and estimate a DCM with measr.
The ROAR-PA data is available in the dcmdata package, and contains responses to `r ncol(roarpa_data) - 1` items from `r fmt_count(nrow(roarpa_data))` respondents.

```{r}
#| label: roarpa-data

roarpa_data
```

In addition to our response data, a DCM also requires a Q-matrix.
A Q-matrix contains one row per item, and one column per attribute (plus an optional column of item identifiers).
A value of 1 indicates that the item measures the attribute, and a value of 0 indicates the item does not measure the attribute.
In our Q-matrix, we can see that the item identifiers in the in the rows (`item`) correspond to the column names of the data.
Additionally, we see that there are three attributes measured by this assessment: `lsm`, `del`, and `fsm`.
These refer to the first sound made (`fsm`), last sound made (`lsm`), and deletion (`del`) elements of phonological awareness.

```{r}
#| label: roarpa-qmatrix

roarpa_qmatrix
```

Our task is to determine which attributes each respondent is proficient on, given their item responses.
For more information on the data set, see `?roarpa` and @roarpa.

## Specify a DCM

A DCM model specification has three primary components: the Q-matrix, a measurement model, and a structural model.
Given these three components, we can create a model specification with `dcm_specify()`.

```{r}
#| label: roarpa-spec

roarpa_spec <- dcm_specify(
  qmatrix = roarpa_qmatrix,
  identifier = "item",
  measurement_model = lcdm(),
  structural_model = unconstrained()
)

roarpa_spec
```

The Q-matrix, as we described, defines which items measure each attribute.
In addition the the Q-matrix itself, we must also tell `dcm_specify()` which column within the Q-matrix contains the item identifiers.
If there is no item identifier, then `identifier` can be left as `NULL` (the default).
In our ROAR-PA specification, we can see that each of our three attributes is measured by 19 items.
The ROAR-PA Q-matrix is a simple structure, meaning that each item measures only one attribute.

At a high level, the measurement model describes how attributes interact with each other on specific items.
If an item measures two attributes, how do we expect a respondent to perform if they possess only one of the attributes?
Are the attributes compensatory, meaning that proficiency on either is sufficient to answer the item correctly, or noncompensatory, and proficiency on both attributes is required in order to provide a correct response?
The choice of measurement model dictates these relationships.

The structural model describes relationships between proficiency on the attributes.
Is proficiency on one attribute independent of proficiency on another, or is proficiency correlated?
It's also possible that some attributes may represent prerequisite knowledge such that respondents must demonstrate proficiency before they can demonstrate proficiency of other attributes.
The structural model is used to define these relationships.

We'll explore both measurement and structural models in more detail in the next sections.

### Measurement models

measr provides functionality for seven DCM measurement models: the six core models identified by @rupp-dcm and a general model that subsumes the other models.
A full description of these models is beyond the scope of what we are covering here.
However, we will provide a high-level overview the types of models and offer referenes for further details on each.

The general DCM supported by the measr is the loglinear cognitive diagnostic model [LCDM\; @lcdm; @lcdm-handbook].
This is the most flexible model that allows each item to have unique interactions between attributes, estimating separate main effects and interaction effects for all possible attribute combinations.
You can think of the LCDM as the "saturated model" that all other DCMs are constrained versions of.
That is, by placing constraints on the LCDM parameters, you can achieve models equivalent to the other core models.

Under the umbrella of the LCDM are the six core DCMs, which generally fall into two categories: non-compensatory (also called conjunctive) and compensatory (disjunctive).
When using a non-compensatory model, attributes function like prerequisites or requirements, and missing an attribute creates a specific deficit that other attributes cannot overcome.
In other words, with non-compensatory models, performance is constrained by the weakest link.
In this category, measr supports the deterministic input, noisy "and" gate model [DINA\; @dina]; the noisy-input, deterministic "and" gate model [NIDA\; @nida]; and the non-compensatory reparameterized unified model [NC-RUM\; @ncrum].
On the other hand, when compensatory models, attributes function like independent skills that accumulate, and having more attributes can partially or fully compensate for missing others.
Thus, performance improves as you gain more attributes.
In the compensatory category, measr supports the deterministic input, noisy "or" gate model [DINO\; @dino]; the noisy-input, deterministic "or" gate model [NIDO\; @nido]; and the compensatory reparameterized unified model [C-RUM\; @crum].

Each of these measurement models can be estimated with measurement by supplying the respective measurement model function, as shown in @tbl-meas-models, to the `measurement_model` argument of `dcm_specify()`.

```{r}
#| label: tbl-meas-models
#| tbl-cap: Measurement models supported by measr
#| echo: false

library(tidyverse)
library(gt)

tibble(
  model = c("LCDM", "DINA", "NIDA", "NC-RUM", "DINO", "NIDO", "C-RUM"),
  description = c(
    "General and flexible, subsumes other models",
    "All attributes must be present",
    "Attributes have multiplicative penalties equal across items",
    "Attributes have multiplicative penalites that vary across items",
    "Any one attribute must be present",
    "Attributes are additive and equal across items",
    "Attributes are additive and vary across items"
  ),
  measr = c(
    "`lcdm()`",
    "`dina()`",
    "`nida()`",
    "`ncrum()`",
    "`dino()`",
    "`nido()`",
    "`crum()`"
  )
) |> 
  gt() |> 
  fmt_markdown(columns = c(measr)) |> 
  cols_width(
    model ~ px(155),
    description ~ px(500),
    measr ~ px(80)
  ) |> 
  tab_row_group(
    label = "Non-compensatory",
    rows = 2:4
  ) |> 
  tab_row_group(
    label = "Compensatory",
    rows = 5:7
  ) |> 
  row_group_order(groups = c(NA, "Non-compensatory", "Compensatory")) |> 
  gt_theme_rdcm() |> 
  tab_style(
    style = list(
      cell_fill(color = msr_colors[2]),
      cell_text(color = "white"),
      cell_borders(weight = px(0))
    ),
    locations = cells_row_groups()
  )
```

### Structural models

measr provides functionality for five structural models.

### Prior distributions



## Estimate a model specification

```{r}
#| label: estimate-roarpa

roarpa_lcdm <- dcm_estimate(
  dcm_spec = roarpa_spec,
  data = roarpa_data,
  identifier = "id",
  method = "optim",
  backend = "rstan",
  file = here::here("start", "fits", "roarpa-lcdm-uncst-rstn")
)
```

### Respondent proficiency estimates

```{r}
#| label: score-roarpa

roarpa_scores <- score(roarpa_lcdm) 
roarpa_scores
```

```{r}
#| label: classify-roarpa
#| message: false

library(tidyverse)

roarpa_scores$attribute_probabilities |> 
  mutate(probability = as.integer(probability > .5)) |> 
  pivot_wider(names_from = attribute, values_from = probability)
```

## Outline

The first article will focus on how to specify a DCM with measr including:

* Choosing a measurement model
* Choosing a structural model
* Choosing prior distributions
* Combining these choices into a DCM specification using `dcm_specify()`

The article will then describe how to estimate the model with MCMC or an optimizer using either `{rstan}` or `{cmdstanr}` with `dcm_estimate()`.

## Session information {#session-info}

```{r}
#| label: "si"
#| echo: false

small_session(pkgs)
```
