---
title: "Specify a diagnostic model"
weight: 1
description: |
  Get started by learning how to specify and estimate a model using measr.
css: ../styles.css
link-citations: true
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

source(here::here("common.R"))
```

```{r}
#| label: load
#| include: false
#| message: false
#| warning: false

library(dcmdata)
library(dcmstan)
library(measr)
pkgs <- c("dcmdata", "measr", "rstan")
```

## Introduction

How do you specify and estimate a diagnostic classification model (DCM) using measr?
In this article, we will walk you through the steps.
We start with data for building the model, learn how to specify DCMs that make different assumptions about the data, and explore how to estimate the model with [*Stan*](https://mc-stan.org).

`r article_req_pkgs(pkgs)`

```{r}
#| label: user-load
#| message: false

library(measr)     # for model specification and estimation

# Helper packages
library(dcmdata)   # for example data set
library(rstan)     # for estimation backend
```

## Rapid Online Assessment of Reading and Phonological Awareness (ROAR-PA) data

Let's use data from the ROAR-PA [@roarpa] to learn how to specify and estimate a DCM with measr.
The ROAR-PA data is available in the dcmdata package, and contains responses to `r ncol(roarpa_data) - 1` items from `r fmt_count(nrow(roarpa_data))` respondents.

```{r}
#| label: roarpa-data

roarpa_data
```

In addition to our response data, a DCM also requires a Q-matrix.
A Q-matrix contains one row per item, and one column per attribute (plus an optional column of item identifiers).
A value of 1 indicates that the item measures the attribute, and a value of 0 indicates the item does not measure the attribute.
In our Q-matrix, we can see that the item identifiers in the in the rows (`item`) correspond to the column names of the data.
Additionally, we see that there are three attributes measured by this assessment: `lsm`, `del`, and `fsm`.
These refer to the first sound made (`fsm`), last sound made (`lsm`), and deletion (`del`) elements of phonological awareness.

```{r}
#| label: roarpa-qmatrix

roarpa_qmatrix
```

Our task is to determine which attributes each respondent is proficient on, given their item responses.
For more information on the data set, see `?roarpa` and @roarpa.

## Specify a DCM

A DCM model specification has three primary components: the Q-matrix, a measurement model, and a structural model.
Given these three components, we can create a model specification with `dcm_specify()`.

```{r}
#| label: roarpa-spec

roarpa_spec <- dcm_specify(
  qmatrix = roarpa_qmatrix,
  identifier = "item",
  measurement_model = lcdm(),
  structural_model = unconstrained()
)

roarpa_spec
```

The Q-matrix, as we described, defines which items measure each attribute.
In addition the the Q-matrix itself, we must also tell `dcm_specify()` which column within the Q-matrix contains the item identifiers.
If there is no item identifier, then `identifier` can be left as `NULL` (the default).
In our ROAR-PA specification, we can see that each of our three attributes is measured by 19 items.
The ROAR-PA Q-matrix is a simple structure, meaning that each item measures only one attribute.

At a high level, the measurement model describes how attributes interact with each other on specific items.
If an item measures two attributes, how do we expect a respondent to perform if they possess only one of the attributes?
Are the attributes compensatory, meaning that proficiency on either is sufficient to answer the item correctly, or noncompensatory, and proficiency on both attributes is required in order to provide a correct response?
The choice of measurement model dictates these relationships.

The structural model describes relationships between proficiency on the attributes.
Is proficiency on one attribute independent of proficiency on another, or is proficiency correlated?
It's also possible that some attributes may represent prerequisite knowledge such that respondents must demonstrate proficiency before they can demonstrate proficiency of other attributes.
The structural model is used to define these relationships.

We'll explore both measurement and structural models in more detail in the next sections.

### Measurement models

measr provides functionality for seven DCM measurement models: the six core models identified by @rupp-dcm and a general model that subsumes the other models.
A full description of these models is beyond the scope of what we are covering here.
However, we will provide a high-level overview the types of models and offer referenes for further details on each.

The general DCM supported by the measr is the loglinear cognitive diagnostic model [LCDM\; @lcdm; @lcdm-handbook].
This is the most flexible model that allows each item to have unique interactions between attributes, estimating separate main effects and interaction effects for all possible attribute combinations.
You can think of the LCDM as the "saturated model" that all other DCMs are constrained versions of.
That is, by placing constraints on the LCDM parameters, you can achieve models equivalent to the other core models.

Under the umbrella of the LCDM are the six core DCMs, which generally fall into two categories: non-compensatory (also called conjunctive) and compensatory (disjunctive).
When using a non-compensatory model, attributes function like prerequisites or requirements, and missing an attribute creates a specific deficit that other attributes cannot overcome.
In other words, with non-compensatory models, performance is constrained by the weakest link.
In this category, measr supports the deterministic input, noisy "and" gate model [DINA\; @dina]; the noisy-input, deterministic "and" gate model [NIDA\; @nida]; and the non-compensatory reparameterized unified model [NC-RUM\; @ncrum].
On the other hand, when compensatory models, attributes function like independent skills that accumulate, and having more attributes can partially or fully compensate for missing others.
Thus, performance improves as you gain more attributes.
In the compensatory category, measr supports the deterministic input, noisy "or" gate model [DINO\; @dino]; the noisy-input, deterministic "or" gate model [NIDO\; @nido]; and the compensatory reparameterized unified model [C-RUM\; @crum].

Each of these measurement models can be estimated with measr by supplying the respective measurement model function, as shown in @tbl-meas-models, to the `measurement_model` argument of `dcm_specify()`.

```{r}
#| label: tbl-meas-models
#| tbl-cap: Measurement models supported by measr
#| echo: false

library(tidyverse)
library(gt)

tibble(
  model = c("LCDM", "DINA", "NIDA", "NC-RUM", "DINO", "NIDO", "C-RUM"),
  description = c(
    "General and flexible, subsumes other models",
    "All attributes must be present",
    "Attributes have multiplicative penalties equal across items",
    "Attributes have multiplicative penalites that vary across items",
    "Any one attribute must be present",
    "Attributes are additive and equal across items",
    "Attributes are additive and vary across items"
  ),
  measr = c(
    "`lcdm()`",
    "`dina()`",
    "`nida()`",
    "`ncrum()`",
    "`dino()`",
    "`nido()`",
    "`crum()`"
  )
) |> 
  gt() |> 
  fmt_markdown(columns = c(measr)) |> 
  cols_width(
    model ~ px(155),
    description ~ px(500),
    measr ~ px(80)
  ) |> 
  tab_row_group(
    label = "Non-compensatory",
    rows = 2:4
  ) |> 
  tab_row_group(
    label = "Compensatory",
    rows = 5:7
  ) |> 
  row_group_order(groups = c(NA, "Non-compensatory", "Compensatory")) |> 
  gt_theme_rdcm() |> 
  tab_style(
    style = list(
      cell_fill(color = msr_colors[2]),
      cell_text(color = "white"),
      cell_borders(weight = px(0))
    ),
    locations = cells_row_groups()
  )
```

### Structural models

measr provides functionality for five structural models.
The structural model describes the joint distribution of attribute profiles in the population.
Different structural models make different assumptions about how attributes relate to each other.

The most general option is the unconstrained model [@rupp-dcm].
This model places no constraints on the relationships between attributes.
Each of the 2^*A*^ possible attribute profiles (where *A* is the number of attributes) has its own freely estimated base rate parameter.
Because all profiles are freely estimated, this is a saturated structural model.

The independent model [@independent] assumes that attributes are completely unrelated.
Proficiency on one attribute tells you nothing about proficiency on another.
Under this model, each attribute has its own proficiency base rate, and the probability of any profile is simply the product of the individual attribute base rates (or their complements for non-proficiency).

The loglinear model [@loglinear] uses a log-linear parameterization with main effects and interactions.
When specifying a loglinear model, we can use the `max_interaction` argument to control the high-level interactions to include.
When `max_interaction` is set to the number of attributes (the default), the loglinear model is equivalent to the unconstrained model.
When `max_interaction = 1`, only main effects are included, which is equivalent to the independent model.
Intermediate values allow you to model some degree of attribute dependence without fully saturating the structural model.

The remaining two structural models incorporate attribute hierarchies.
In these models, proficiency on some attributes may be a prerequisite for proficiency on others.
The hierarchical DCM [HDCM\; @hdcm] enforces strict attribute prerequisites.
Attribute profiles that violate the specified hierarchy are excluded entirely from the model, meaning their base rates are fixed to zero.
In contrast, the Bayesian network model [@bayesnet] implements a softer version of the hierarchy.
All attribute profiles remain possible, but profiles that are inconsistent with the hierarchy are estimated to be less likely.
Both models require a `hierarchy` argument that defines the attribute relationships using dagitty-style syntax, such as `"att1 -> att2 -> att3"`.
For more details on specifying attribute hierarchies, see the [Define Attribute Relationships](/start/hierarchies/) article.

Each of these structural models can be estimated with measr by supplying the respective structural model function, as shown in @tbl-strc-models, to the `structural_model` argument of `dcm_specify()`.

```{r}
#| label: tbl-strc-models
#| tbl-cap: Structural models supported by measr
#| echo: false

library(tidyverse)
library(gt)

tibble(
  model = c("Unconstrained", "Independent", "Loglinear", "HDCM", "BayesNet"),
  description = c(
    "General and flexible, subsumes other models",
    "Attributes are independent of each other",
    "Can be constrained to only include certain interaction levels",
    "Hard constraints on profiles based on attribute dependencies",
    "Soft constraints on profiles based on attribute dependencies"
  ),
  measr = c(
    "`unconstrained()`",
    "`independent()`",
    "`loglinear()`",
    "`hdcm()`",
    "`bayesnet()`"
  )
) |> 
  gt() |> 
  fmt_markdown(columns = c(measr)) |> 
  cols_width(
    model ~ px(130),
    description ~ px(475),
    measr ~ px(150)
  ) |> 
  gt_theme_rdcm()
```

### Prior distributions

A final aspect of the DCM specification that we have not yet talked about is the definition of the model priors.
Take another look at our specifciation object.
We can see that there are prior distribution defined for each type of parameter in our model.

```{r}
#| label: print-roarpa

roarpa_spec
```

Every parameter in a DCM specification is assigned a prior distribution that encodes our beliefs about plausible parameter values before observing any data.
measr provides sensible defaults, but you can also customize priors to reflect domain knowledge or to implement more informative constraints.

To view the default priors for a given measurement and structural model combination, use `default_dcm_priors()`.
In our example, we have specified an LCDM with an unconstrained structural model.
Plugging those two components in, we see the same priors that we saw in our specification object.

```{r}
#| label: default-priors

default_dcm_priors(
  measurement_model = lcdm(),
  structural_model = unconstrained()
)
```

However, different choices of measurement and structural models will result in different parameters being included, and therefore different prior distributions.
For example, specifying a DINA measurement model with and independent structural model has a completely different set of parameters.

```{r}
#| label: dina-params

default_dcm_priors(
  measurement_model = dina(),
  structural_model = independent()
)
```

You can see which parameter types and specific coefficients are available for your model using `get_parameters()`.

```{r}
#| label: get-params

get_parameters(lcdm(), qmatrix = roarpa_qmatrix, identifier = "item")
```

To customize priors, use the `prior()` function.
The `type` argument specifies which parameter type the prior applies to, and the optional `coefficient` argument can target a specific parameter within that type.
Custom priors can be passed to `dcm_specify()` via the `priors` argument.
Any parameter types not covered by a custom prior will retain their default values.

```{r}
#| label: custom-priors

my_priors <- c(
  prior(normal(0, 1), type = "intercept"),
  prior(lognormal(0, 0.5), type = "maineffect", coefficient = "l1_13")
)

dcm_specify(
  qmatrix = roarpa_qmatrix,
  identifier = "item",
  measurement_model = lcdm(),
  structural_model = unconstrained(),
  priors = my_priors
)
```

## Estimate a model specification

Once we have a model specification, we can estimate it using `dcm_estimate()`.
This function takes a specification object (created by `dcm_specify()`), along with the response data and the name of the column in the data that contains respondent identifiers.

The `method` argument controls how the model is estimated.
Options include `"optim"` for point estimation using Stan's optimizer, `"mcmc"` for full Markov chain Monte Carlo sampling, `"variational"` for variational inference, and `"pathfinder"` (available only when using the cmdstanr backend).
Full MCMC provides the most complete picture of the posterior distribution, but takes the longest to run.
The optimizer is the fastest option and is useful for quick analyses, but does not provide a full posterior distribution.

The `backend` argument specifies which Stan interface to use for estimation: `"rstan"` or `"cmdstanr"` to use the `{rstan}` or `{cmdstanr}` package, respetively.
The `file` argument allows you to save the estimated model to disk so that it does not need to be re-estimated if you re-run the script.
Any additional arguments are passed directly to the backend's estimation function (e.g., `chains`, `iter`, and `warmup` for MCMC estimation when using the `"rstan"` backend).

For this example, we use the optimizer with rstan, which provides fast point estimates of the model parameters.

```{r}
#| label: estimate-roarpa

roarpa_lcdm <- dcm_estimate(
  dcm_spec = roarpa_spec,
  data = roarpa_data,
  identifier = "id",
  method = "optim",
  backend = "rstan",
  file = here::here("start", "fits", "roarpa-lcdm-uncst-rstn")
)
```

### Respondent proficiency estimates

After estimating a model, we typically want to know which attributes each respondent has mastered.
The `score()` function calculates respondent proficiency estimates from a fitted model.
It returns a list with two elements: `class_probabilities`, which contains the probability that each respondent belongs to each possible attribute profile, and `attribute_probabilities`, which contains the marginal probability that each respondent is proficient on each individual attribute.

```{r}
#| label: score-roarpa

roarpa_scores <- score(roarpa_lcdm)
roarpa_scores
```

In practice, we often want to convert these probabilities into binary proficiency classifications.
A common approach is to use a threshold of .5, classifying a respondent as proficient on an attribute if their estimated probability of proficiency exceeds .5.
The choice of threshold matters and can be adjusted based on the intended use of the results.

```{r}
#| label: classify-roarpa
#| message: false

library(tidyverse)

roarpa_scores$attribute_probabilities |> 
  mutate(probability = as.integer(probability > .5)) |> 
  pivot_wider(names_from = attribute, values_from = probability)
```

## Wrapping up

We now have an estimate of proficiency for each respondent on each of the attribute measured by the ROAR-PA.
However, before we report these result it's important to evaluate the quality of the model.
We need to ensure that the model fits well and provides accurate classifications.
That is the focus of the [Evaluate Model Performance](/start/evaluate/) article.

## Session information {#session-info}

```{r}
#| label: "si"
#| echo: false

small_session(pkgs)
```
